---
title: "An an descriptive analysis of the results of Enem"
author: "Maria Sales"
date: "02/12/2021"
output: pdf_document
---
This project is an a descriptive analyses of results of the Enem. The goal is 
create similar groupsbased on the grades of students in the three areas of the 
examen.Based on that groups, it was perfomed an descriptive analysis to 
understand the main characteristicas ofeach group. 
```{r}
#Loading the packages
library(tidyverse) #pacote para manipulacao de dados
library(cluster) #algoritmo de cluster
library(dendextend) #compara dendogramas
library(factoextra) #algoritmo de cluster e visualizacao
library(fpc) #algoritmo de cluster e visualizacao
library(gridExtra) #para a funcao grid arrange
library(readxl)
library(readr)
```

Reading the data set and checking the dimension of the data
```{r}
df<- read_csv("C:/Users/jeova/Downloads/microdados_com_proficiencia_sample.csv")
View(df)
dim(df)
```

The data set was filtered in to order excluder null values. These values was 
removed because theycreate a bias in the analyses. That is why the students who
did not attend of the the days of the examem received a zero score and this score 
does not reflect the perfomance of the student. 
```{r}
#Excluding outliers
df <- filter(df,NU_NOTA_CH != 0 & NU_NOTA_CN != 0)
dim(df)
```
The clustering was carried out using the k-mean algortimh and the hierarchical 
method. First of all, the data set was divided in order to keep just the 
variables which would be used in the clusterization.
```{r}
"Creating a new data set"
df_1<- data.frame(df$NU_INSCRICAO,df$NU_NOTA_CH,df$NU_NOTA_CN)
View(df_1)

# setting the name of the rows
rownames(df_1)<- df_1[,1]
df_1 <- df_1[,-1]
View(df_1)
```

The hierarquical method was used to establish the number of groups to be used in
the k-means method. How the data set is too big with 50000 rows. It is not 
appropriate to use just the hierarquical method for the clusterizion. 
The k- mean method is more suitable, but is necessary first feed the algorith 
with the numbers of clusters   

For that, it was selected randomly a numbers of rows to create a smaller data 
set and so on perfomerthe hierarquical method. 
```{r}
# Selecting randomless a numbers of rows to create a new dataframe
df_2 <- sample_n(df_1, size = 3000, replace = FALSE)
view(df_2)

#Calculating the matrix distance
distance_2 <- dist(df_2, method = 'euclidean')

#Clustering using the average method
C_3<-hclust(distance_2, method='average')
```

The dendogram was plotted to see the formation of the groups
```{r}
#Plotting the dendogram
plot(C_3, cex = 0.6, hang = -1)
```

The elbow method was used to establish the number of groups that will be used in the k-mean method, It was noticed that 4 groups seems like a ideal numbers to proceed the analysis. 
```{r}
# Using the elbow method to establish the number of groups
fviz_nbclust(df_2, FUN = hcut, method = 'wss')
```

The k-means algorthim was perfomend using 3,4,5 and 6 centers to seeing the how the groups changes
according to the number of the groups. As said before, 4 groups seems a suitable number for the analysis.
```{r}
C_k3 <- kmeans(df_1, centers = 3)
C_k4 <- kmeans(df_1, centers = 4)
C_k5 <- kmeans(df_1, centers = 5)
C_k6 <- kmeans(df_1, centers = 6)

#Transforming the cluster in a dataset
C_k4_df <- data.frame(C_k4$cluster)
View(C_k4_df)
```

The clusterization with all the suggested groups can be seen in the graph below.
```{r}
# Visualizing the clusters
G_k3 <- fviz_cluster(C_k3, geom = 'point', data = df_1, main = 'Cluster_k3')
G_k4 <- fviz_cluster(C_k4, geom = 'point', data = df_1, main = 'Cluster_k4')
G_k5 <- fviz_cluster(C_k5, geom = 'point', data = df_1, main = 'Cluster_k5')
G_k6 <- fviz_cluster(C_k6, geom = 'point', data = df_1, main = 'cluster_k6')

# Printing the graphs together

grid.arrange(G_k3,G_k4,G_k5,G_k6, nrow = 2 )

```


An 2D graph was plotted considering 4 groups. 
```{r}
#plotting a 2D graph with ggplot
ggplot(data = df_1) + 
geom_point(aes( x = df.NU_NOTA_CH,
                y = df.NU_NOTA_CN,
               color = as.factor(C_k4_df$C_k4.cluster)))

```

Given that it was chosen 4 groups for the clusterization, it was created a dataframe with the
main characteristics of each group. 
```{r}
# Binding the dataframes
df_fim <- cbind(df, C_k4_df)

#Categorizing the groups created by kmeans

Enem_description <- df_fim %>%
 group_by(C_k4.cluster)%>%
 summarise(n = n(), 
 mean_nota_CH = mean(NU_NOTA_CH, na.rm = TRUE),
 mean_nota_CN = mean(NU_NOTA_CN, na.rm= TRUE),
 mean_nota_LC = mean(NU_NOTA_LC,na.rm= TRUE ),
 mean = (mean_nota_CH+mean_nota_CN+mean_nota_LC)/3,
 sexo_Feminino = table(factor(TP_SEXO),exclude = ('M'), useNA = 'no'),
 sexo_Masculino = table(factor(TP_SEXO),exclude = ('F'),useNA = 'no'),
 raca_n_declarado = table(factor(TP_COR_RACA), exclude = c(1,2,3,4,5),useNA = 'no'),
 raca_branca = table(factor(TP_COR_RACA), exclude = c(0,2,3,4,5),useNA = 'no'),
 raca_preta = table(factor(TP_COR_RACA), exclude = c(0,1,3,4,5),useNA = 'no'),
 raca_parda = table(factor(TP_COR_RACA), exclude = c(0,1,2,4,5),useNA = 'no'),
 raca_amarela = table(factor(TP_COR_RACA), exclude = c(0,1,2,3,5),useNA = 'no'),
 raca_indigena = table(factor(TP_COR_RACA), exclude = c(0,1,2,3,4),useNA = 'no'),
 escola_n_respondeu = table(factor(TP_ESCOLA), exclude = c(2,3,4),useNA = 'no'),
 escola_publica = table(factor(TP_ESCOLA), exclude = c(1,3,4),useNA = 'no'),
 escola_privada = table(factor(TP_ESCOLA), exclude = c(1,2,4),useNA = 'no'))

Enem_description <- relocate(Enem_description, mean, .after = n)

View(Enem_description)

```

To better analyse the characteristics of the groups, the results were transformed 
to proportion
```{r}
Enem_description_pct <-  Enem_description %>%
  select(everything(), -c(C_k4.cluster, n, mean, mean_nota_CH,
  mean_nota_CN, mean_nota_LC))%>%
  as.matrix()%>%
  prop.table(margin = 2)%>%
  data.frame()
  
Enem_description_pct <- cbind(Enem_description[,1:3],
                            Enem_description_pct)

View(Enem_description_pct)
```

The following code represents the descriptive analyses of the cluster considering
the familiar income of the students.
```{r}
# Description analysis of the cluster by income
Enem_description_renda <- df_fim %>%
group_by(C_k4.cluster)%>%
summarise(n = n(), 
mean_nota_CH = mean(NU_NOTA_CH, na.rm = TRUE),
mean_nota_CN = mean(NU_NOTA_CN, na.rm= TRUE),
mean_nota_LC = mean(NU_NOTA_LC,na.rm= TRUE ),
mean = (mean_nota_CH+mean_nota_CN+mean_nota_LC)/3,
renda_A = table(factor(Q006),exclude = c('A','C','D','E','F','G','H','I','J','K','L','M','N','O','P','Q')),
renda_B = table(factor(Q006),exclude = c('B','A','D','E','F','G','H','I','J','K','L','M','N','O','P','Q')),
renda_C = table(factor(Q006),exclude = c('B','C','A','E','F','G','H','I','J','K','L','M','N','O','P','Q')),
renda_D = table(factor(Q006),exclude = c('B','C','D','A','F','G','H','I','J','K','L','M','N','O','P','Q')),
renda_E = table(factor(Q006),exclude = c('B','C','D','E','A','G','H','I','J','K','L','M','N','O','P','Q')),
renda_F = table(factor(Q006),exclude = c('B','C','D','E','F','A','H','I','J','K','L','M','N','O','P','Q')),
renda_G = table(factor(Q006),exclude = c('B','C','D','E','F','G','A','I','J','K','L','M','N','O','P','Q')),
renda_H = table(factor(Q006),exclude = c('B','C','D','E','F','G','H','A','J','K','L','M','N','O','P','Q')),
renda_I = table(factor(Q006),exclude = c('B','C','D','E','F','G','H','I','A','K','L','M','N','O','P','Q')),
renda_J = table(factor(Q006),exclude = c('B','C','D','E','F','G','H','I','J','A','L','M','N','O','P','Q')),
renda_L = table(factor(Q006),exclude = c('B','C','D','E','F','G','H','I','J','K','A','M','N','O','P','Q')),
renda_M = table(factor(Q006),exclude = c('B','C','D','E','F','G','H','I','J','K','L','A','N','O','P','Q')),
renda_O = table(factor(Q006),exclude = c('B','C','D','E','F','G','H','I','J','K','L','M','A','O','P','Q')),
renda_P = table(factor(Q006),exclude = c('B','C','D','E','F','G','H','I','J','K','L','M','N','A','P','Q')),
renda_Q = table(factor(Q006),exclude = c('B','C','D','E','F','G','H','I','J','K','L','M','N','O','A','Q')))

Enem_description_renda <- relocate(Enem_description_renda, mean, .after = n)

View(Enem_description_renda)
```

In the same way, the results were transformed to proportions to better understand 
the influence of the familiar income in the score of each group
```{r}
Enem_description_renda_pct <-  Enem_description_renda %>%
  select(everything(), -c(C_k4.cluster, n, mean, mean_nota_CH,
                          mean_nota_CN, mean_nota_LC))%>%
  as.matrix()%>%
  prop.table(margin = 2)%>%
  data.frame()

Enem_description_renda_pct <- cbind(Enem_description_renda[,1:3],
                              Enem_description_renda_pct)

View(Enem_description_renda_pct)

```

`




